---
title: One-sample *t*-test in R
subtitle: Cheatsheet
date: today
lightbox: true
---

```{r}
#| include: false
library(dplyr)
library(tidyr)
library(gt)
possums <- readxl::read_excel("possum_bw.xlsx")
```

::: {.grid}
::: {.g-col-6}
{{< include _license.qmd >}}

::: {.callout-note collapse="true"}
## Assumed knowledge
- You know how to install and load packages in R.
- You know how to import data into R.
- You recognise data frames and vectors.
:::
:::
::: {.g-col-6}
{{< include _tidy.qmd >}}

::: {.callout-important}
## Data
For this cheatsheet we will use data from the possums dataset used in [BIOL2022](https://www.sydney.edu.au/units/BIOL2022) labs.

{{< downloadthis possums.xlsx dname="possum" label="possum.xlsx" type="primary" >}}
:::
:::
:::

## About 
The one-sample *t*-test is used to determine whether the mean of a single sample $y$ is significantly different from a known or hypothesised population mean ($\mu$). **Examples**:

- Is the mean weight of canned tuna significantly different from what was stated on the label (400 g)?
- Is the mean height of a sample of male students significantly different from the national average height (175.6 cm)? 
- Is the mean number of kittens in a litter significantly different from 4?

## Modelling

> Is the mean **body weight** of possums (`BW`) significantly different from 3.5 kg?

The [**simplified model**]{style="color:olive"} for the mathematically-adverse individual is $$\color{olive}\text{body weight} \sim 3.5$$ which translates to "the body weight of possums is *around* 3.5 kg".
The [**statistical model**]{style="color:red"} is $$\color{red}\text{body weight} = \beta_0 + \epsilon$$ where $\beta_0$ is the hypothesised population mean and $\epsilon$ is the error term.

### Preparing the data
Extract **only** the variable of interest from the dataset using `select()` from the `dplyr` package – `BW`. Assign the variable to a new object -- `bw` in this case.

```{r}
library(dplyr)
library(readxl)
possums <- read_excel("possums.xlsx", sheet = 2) # import
bw <- select(possums, BW) # select variable
```

Your own data should be in a similar format.

## Analytical approaches
The traditional approach to the one-sample *t*-test is to use the `t.test()` function in R, while the modern approach is to use a general linear model (GLM) with the `lm()` or `glm()` functions.

::: panel-tabset

## `t.test()` function

### Methods reporting

> A one-sample *t*-test was used to determine whether the mean body weight of possums was significantly different from 3.5 kg. This was computed using the `t.test()` function in R version 4.4.0 (R Core Team, 2024).

### Perform the analysis
```{r}
#| results: hide
t.test(bw, mu = 3.5)
```

### Check assumption(s)
#### Normality
Any combination of one or more of the following checks can be used to assess normality:

- **Histogram**: `hist(bw$BW)`
- **Q-Q plot**: `qqnorm(bw$BW)`
- **Shapiro-Wilk test**: `shapiro.test(bw$BW)`

Include the appropriate description in your methods section.

> The normality of body weight was assessed using [insert method(s)].

### How to report results
> The mean body weight of possums was significantly different from 3.5 kg (t~19~ = -10.3, 95% CI [2.3, 2.7], p < 0.001).

## `lm()` function

### Methods reporting
> A general linear model was used to determine whether the mean body weight of possums was significantly different from 3.5 kg. This was computed using the `lm()` function in R version 4.4.0 (R Core Team, 2024).

### Perform the analysis

For a one-sample *t*-test, the formula needs to be specified as `y - µ ~ 1` where `y` is the variable of interest and µ is the hypothesised value that is being tested. The `1` indicates that the model has an intercept only i.e. we are testing whether the mean difference is significantly different from 0.

```{r}
#| results: hide
fit <- lm((BW - 3.5) ~ 1, data = bw)
summary(fit)
```

### Check assumption(s)
#### Normality

With a GLM, normality can be assessed using the residuals of the model. The following checks can be used:

- **Histogram**: `hist(residuals(fit))`
- **Q-Q plot**: `qqnorm(residuals(fit))`
- **Shapiro-Wilk test**: `shapiro.test(residuals(fit))`

### How to report results

> There is evidence to suggest that the mean body weight of possums was significantly different from 3.5 kg (GLM, t~19~ = -10.3, p < 0.001).

:::

## Exercise(s)

Download the penguins dataset (from below if you are reading this in HTML), or load the dataset from the `palmerpenguins` package. Perform a one-sample *t*-test to determine whether the mean flipper length of penguins is significantly different from 200 mm. 

{{< downloadthis penguins.csv dname="penguins" label="penguins.csv" type="primary" >}}



